---
title: "LASSO regression and uncertainty estimates for sulfur capacity"
subtitle: "Supplementary notebook for Gorojovsky and Wood (2026)"
author: "L. R. Gorojovsky"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    df-print: paged
execute:
  echo: true
  warning: true
  message: false
---

This notebook accompanies the supplementary R project for:

**Gorojovsky, L. R. and Wood, B. J. (2026). Solubility and speciation of sulfur in silicate melts under crustal conditions. Earth and Planetary Science Letters.**

It reproduces the LASSO regression workflow used to fit sulfide and sulfate capacity models, estimate prediction intervals, and propagate uncertainty into an example sulfur-speciation calculation. The standalone R script, `S2_LASSO_regression.R`, remains the canonical script version of the same workflow.

# Setup

Install the required packages before running the notebook. Most packages are available from CRAN:

```{r}
#| eval: false
install.packages(c(
  "readxl",
  "glmnet",
  "tibble",
  "ggplot2",
  "dplyr",
  "patchwork",
  "ggprism",
  "scales"
))
```

The `conformalInference` package is installed from GitHub:

```{r}
#| eval: false
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

devtools::install_github("ryantibs/conformal", subdir = "conformalInference")
```

Load the packages used in the analysis.

```{r}
required_packages <- c(
  "readxl",
  "glmnet",
  "tibble",
  "ggplot2",
  "dplyr",
  "patchwork",
  "ggprism",
  "scales",
  "conformalInference"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install the following package(s) before running this notebook: ",
    paste(missing_packages, collapse = ", ")
  )
}

invisible(lapply(required_packages, library, character.only = TRUE))
```

# Data Import

The calibration data are read from `S1_Calibration_data_and_fitting_results.xlsx`. The fitted model form is:

`logCS = A0 + B/T + sum(XA/T)`

```{r}
calibration_workbook <- "S1_Calibration_data_and_fitting_results.xlsx"

if (!file.exists(calibration_workbook)) {
  stop("Could not find ", calibration_workbook, " in the project directory.")
}

df_S6 <- read_excel(
  calibration_workbook,
  sheet = "S2_logCS6_data_for_fitting",
  skip = 1
)

df_S2 <- read_excel(
  calibration_workbook,
  sheet = "S1_logCS2_data_for_fitting",
  skip = 1
)

tibble(
  dataset = c("Sulfate capacity, logCS6+", "Sulfide capacity, logCS2-"),
  rows = c(nrow(df_S6), nrow(df_S2)),
  columns = c(ncol(df_S6), ncol(df_S2))
)
```

Build the design matrices used by `glmnet`. The `Study` and `T` columns are excluded from the model matrices.

```{r}
x1 <- model.matrix(logCS6 ~ . - 1 - Study - T, df_S6)
y1 <- df_S6$logCS6

x2 <- model.matrix(logCS2 ~ . - 1 - Study - T, df_S2)
y2 <- df_S2$logCS2

tibble(
  model = c("logCS6+", "logCS2-"),
  observations = c(nrow(x1), nrow(x2)),
  predictors = c(ncol(x1), ncol(x2))
)
```

# Repeated Cross-Validation

The models are fit with LASSO regression using `glmnet`. Repeated Monte Carlo cross-validation summarizes the selected `lambda.1se` value and the distribution of selected coefficients.

The publication workflow uses `B_reps <- 100` and `nfolds <- 20`. For a quick local smoke test, reduce `B_reps`, then restore it before reproducing the supplementary results.

```{r}
set.seed(80085)

B_reps <- 100
nfolds <- 20

run_mc_cv <- function(x, y, B = 100, nfolds = 20) {
  lambda_1se_vals <- numeric(B)
  rmse_1se_vals <- numeric(B)
  coeff_list <- vector("list", B)

  for (b in seq_len(B)) {
    cv <- cv.glmnet(x, y, alpha = 1, nfolds = nfolds, type.measure = "mse")

    lambda_1se_vals[b] <- cv$lambda.1se
    rmse_1se_vals[b] <- sqrt(cv$cvm[cv$index["1se", 1]])

    beta_1se <- coef(cv, s = "lambda.1se")
    coeff_list[[b]] <- tibble(
      term = beta_1se@Dimnames[[1]][beta_1se@i + 1],
      value = beta_1se@x
    )
  }

  coef_tbl <- bind_rows(coeff_list, .id = "iter") |>
    group_by(term) |>
    summarise(
      mean = mean(value),
      sd = sd(value),
      n_non0 = n(),
      .groups = "drop"
    ) |>
    mutate(select_freq = n_non0 / B)

  list(
    lambda_1se_mean = mean(lambda_1se_vals),
    lambda_1se_sd = sd(lambda_1se_vals),
    rmse_1se_mean = mean(rmse_1se_vals),
    coef_tbl = coef_tbl
  )
}

cv_S6 <- run_mc_cv(x1, y1, B = B_reps, nfolds = nfolds)
cv_S2 <- run_mc_cv(x2, y2, B = B_reps, nfolds = nfolds)

cv_summary <- tibble(
  model = c("Sulfate capacity, logCS6+", "Sulfide capacity, logCS2-"),
  lambda_1se_mean = c(cv_S6$lambda_1se_mean, cv_S2$lambda_1se_mean),
  lambda_1se_sd = c(cv_S6$lambda_1se_sd, cv_S2$lambda_1se_sd),
  mean_rmse = c(cv_S6$rmse_1se_mean, cv_S2$rmse_1se_mean)
)

cv_summary
```

Coefficient selection summaries from the repeated cross-validation.

```{r}
cv_S6$coef_tbl |>
  arrange(desc(select_freq), term)
```

```{r}
cv_S2$coef_tbl |>
  arrange(desc(select_freq), term)
```

# Final LASSO Fits

The final models are fit on all calibration data using the mean `lambda.1se` values from the repeated cross-validation.

```{r}
fit_S6 <- glmnet(x1, y1, alpha = 1, lambda = cv_S6$lambda_1se_mean)
beta_S6 <- as.matrix(coef(fit_S6))
beta_S6_sparse <- beta_S6[beta_S6[, 1] != 0, , drop = FALSE]

fit_S2 <- glmnet(x2, y2, alpha = 1, lambda = cv_S2$lambda_1se_mean)
beta_S2 <- as.matrix(coef(fit_S2))
beta_S2_sparse <- beta_S2[beta_S2[, 1] != 0, , drop = FALSE]

as_tibble(beta_S6_sparse, rownames = "term") |>
  rename(coefficient = s0)
```

```{r}
as_tibble(beta_S2_sparse, rownames = "term") |>
  rename(coefficient = s0)
```

# Prediction Intervals

Split-conformal prediction is used to estimate 95% prediction intervals with the LASSO penalty fixed from the repeated cross-validation.

```{r}
train_S6 <- function(x, y) {
  glmnet(
    x,
    y,
    alpha = 1,
    lambda = cv_S6$lambda_1se_mean,
    standardize = TRUE
  )
}

predict_S6 <- function(fit, newx) {
  as.numeric(predict(fit, newx, s = cv_S6$lambda_1se_mean))
}

conf_S6 <- conformal.pred.split(
  x = x1,
  y = y1,
  x0 = x1,
  train.fun = train_S6,
  predict.fun = predict_S6,
  alpha = 0.05
)

pred_S6 <- df_S6 |>
  mutate(pred = conf_S6$pred, lo95 = conf_S6$lo, hi95 = conf_S6$up)
```

```{r}
train_S2 <- function(x, y) {
  glmnet(
    x,
    y,
    alpha = 1,
    lambda = cv_S2$lambda_1se_mean,
    standardize = TRUE
  )
}

predict_S2 <- function(fit, newx) {
  as.numeric(predict(fit, newx, s = cv_S2$lambda_1se_mean))
}

conf_S2 <- conformal.pred.split(
  x = x2,
  y = y2,
  x0 = x2,
  train.fun = train_S2,
  predict.fun = predict_S2,
  alpha = 0.05
)

pred_S2 <- df_S2 |>
  mutate(pred = conf_S2$pred, lo95 = conf_S2$lo, hi95 = conf_S2$up)
```

# Measured Versus Predicted

The following plot reproduces the measured-vs-predicted comparison for the sulfide and sulfate capacity models.

```{r}
#| fig-width: 9
#| fig-height: 4.8
study_levels <- c("OM", "BW", "GW")
temp_range <- c(1000, 1500)

common_fill <- scale_fill_viridis_c(
  option = "turbo",
  name = "T (deg C)",
  limits = temp_range,
  breaks = seq(1050, 2100, 200),
  oob = scales::squish
)

res_S6 <- pred_S6$logCS6 - pred_S6$pred
rmse_S6 <- sqrt(mean(res_S6^2))
delta_S6 <- 1.96 * rmse_S6

pred_S6 <- pred_S6 |>
  arrange(factor(Study, levels = study_levels)) |>
  mutate(Study = factor(Study, levels = study_levels))

pS6 <- ggplot(pred_S6, aes(x = logCS6, y = pred, shape = Study, fill = T - 273)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_abline(slope = 1, intercept = delta_S6, linetype = "dotted") +
  geom_abline(slope = 1, intercept = -delta_S6, linetype = "dotted") +
  geom_errorbar(aes(ymin = lo95, ymax = hi95), width = 0, linewidth = 0.2) +
  geom_point(size = 2) +
  scale_shape_manual(values = c(22, 23, 21)) +
  common_fill +
  labs(x = "Measured logCS6+", y = "Predicted logCS6+") +
  coord_cartesian(clip = "off") +
  annotation_ticks(sides = "bl", type = "both", outside = TRUE) +
  scale_x_continuous(limits = c(2, 13), breaks = seq(2, 13, by = 2)) +
  scale_y_continuous(limits = c(2, 13), breaks = seq(2, 13, by = 2)) +
  theme_minimal() +
  theme(
    legend.key.size = unit(10, "point"),
    aspect.ratio = 0.8,
    panel.grid.major.y = element_line(color = "grey", linewidth = 0.3, linetype = 2),
    panel.background = element_rect(colour = "black", fill = NA)
  )

res_S2 <- pred_S2$logCS2 - pred_S2$pred
rmse_S2 <- sqrt(mean(res_S2^2))
delta_S2 <- 1.96 * rmse_S2

pred_S2 <- pred_S2 |>
  arrange(factor(Study, levels = study_levels)) |>
  mutate(Study = factor(Study, levels = study_levels))

pS2 <- ggplot(pred_S2, aes(x = logCS2, y = pred, shape = Study, fill = T - 273)) +
  geom_errorbar(aes(ymin = lo95, ymax = hi95), width = 0, linewidth = 0.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_abline(slope = 1, intercept = delta_S2, linetype = "dotted") +
  geom_abline(slope = 1, intercept = -delta_S2, linetype = "dotted") +
  geom_point(size = 2) +
  scale_shape_manual(values = c(22, 23, 21)) +
  common_fill +
  labs(x = "Measured logCS2-", y = "Predicted logCS2-") +
  coord_cartesian(clip = "off") +
  annotation_ticks(sides = "bl", type = "both", outside = TRUE) +
  scale_x_continuous(limits = c(-8, -2), breaks = seq(-8, -2, by = 1)) +
  scale_y_continuous(limits = c(-8, -2), breaks = seq(-8, -2, by = 1)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    aspect.ratio = 0.8,
    panel.grid.major.y = element_line(color = "grey", linewidth = 0.3, linetype = 2),
    panel.background = element_rect(colour = "black", fill = NA)
  )

pS2 + pS6 + plot_layout(ncol = 2)
```

# Monte Carlo Propagation Example

This section propagates uncertainty in `logCS6+`, `logCS2-`, and `logfO2` into an example `S6+/Stot` calculation. The input values below are an example and should be modified for other compositions or conditions.

```{r}
#| fig-width: 5.5
#| fig-height: 4.5
mu6 <- 10.824
mu2 <- -6.226
se6 <- 0.55 / 1.96
se2 <- 0.35 / 1.96
se_fO2 <- 0.01

set.seed(80085)
R <- 50000
draw6 <- rnorm(R, mu6, se6)
draw2 <- rnorm(R, mu2, se2)

fO2_vec <- seq(-11, -5, by = 0.1)

out <- lapply(fO2_vec, function(lgfO2) {
  L_draw <- draw6 - draw2 + 2 * (lgfO2 + rnorm(1, 0, se_fO2))
  frac <- 1 - 1 / (1 + 10^L_draw)

  tibble(
    point = 1 - 1 / (1 + 10^(mu6 - mu2 + 2 * lgfO2)),
    lo95 = quantile(frac, 0.025),
    hi95 = quantile(frac, 0.975)
  )
})

grid_df <- bind_rows(out) |>
  mutate(logfO2 = fO2_vec)

ggplot(grid_df, aes(x = logfO2 - (-25096.3 / 1323 + 8.735), y = point)) +
  geom_ribbon(aes(ymin = lo95, ymax = hi95), fill = "purple", alpha = 0.3) +
  geom_line(linewidth = 1) +
  coord_cartesian(clip = "off") +
  annotation_ticks(sides = "bl", type = "both", outside = TRUE) +
  scale_x_continuous(limits = c(-1, 4), breaks = seq(-1, 4, by = 1)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(x = expression(logfO[2] * Delta * FMQ), y = "S6+ fraction") +
  theme_minimal() +
  theme(
    aspect.ratio = 0.8,
    panel.grid.major.y = element_line(color = "grey", linewidth = 0.3, linetype = 2),
    panel.background = element_rect(colour = "black", fill = NA)
  )
```

# Session Information

Record the R and package versions used to run the notebook.

```{r}
sessionInfo()
```
